Extreme heterogeneity of influenza virus infection in single cells

### 2018-01-14 paper/ResponseToReviews.tex

- As the variability of the viral RNA load appears to be the central result of the manuscript, it would be useful to:

    a. Employ a statistic to quantify the variability (e.g. entropy or Gini index)

    b. Use simple models to illustrate how much variability can be accounted for by simple effects, such as expected Poisson co-infection frequency, or the likelihood of attaining full complement of RNP genes.

{\color{black}
The idea of quantifying the variability in viral mRNA across infected cells is a good one.
We have done this using the Gini coefficient.
The results are in Figure 4--Figure supplement 3, and are referred to in the text related to that figure.
As a fun point of comparison, the Gini coefficients for the variability of viral mRNA across cells ($\ge$0.64 for all samples) exceed the Gini coefficient for the unevenness of income distribution in the United States.

We have also simulated the expected distribution under a simple model where the heterogeneity is due only to Poisson co-infection and whether or not a cell has the full complement of RNP genes.
This simulation and its results are in Figure 5--source data 2. 
The simulation gives a Gini coefficient as large as the actual one, but the shape of the simulated distribution is essentially bimodal, whereas the actual distribution is much more continuous and broad.
Therefore, there are also other sources of heterogeneity that broaden the distribution beyond what is due only to RNP presence / absence and co-infection. 
}

### 2018-01-14 paper/paper.tex
added simple simulation model requested by reviewers

\begin{figure}[t!]
\centerline{\includegraphics[width=0.9\linewidth]{figures/p_flu_burden_flu_gene_merge.pdf}}
\caption{\label{fig:fluburdenbyflugene}
The absence of viral genes explains some of the variability in the amount of viral mRNA per cell.
{\bf (A)} 
The normalized expression of each viral gene as a function of the total fraction of mRNA in each infected cell derived from virus, taken over all time points.
Cells with high viral burden always express all RNP genes, but some cells with high viral burden lack each of the other genes.
Plots for individual samples are in \FIGSUPP[fluburdenbyflugene]{fluburdensamplebreakdown}, and a plot that excludes known coinfected cells is in \FIGSUPP[fluburdenbyflugene]{fluburden_nocoinfection}.
{\bf (B)}
Box and whisker plots showing the per-cell viral burden among cells with $>$0.5\% of their mRNA from virus, binned by whether or not the cells express each gene.
A Wilcoxon signed-rank test was used to test the null hypothesis that absence of each gene does not affect viral burden: **** = $P < 10^{-4}$, *** = $P < 10^{-3}$,  * = $P < 0.05$, ns = not significant.
The trends are similar if we look only at the 10-hour sample (\FIGSUPP[fluburdenbyflugene]{10hrfluburdenbyflugene}) or exclude known co-infected cells (\FIGSUPP[fluburdenbyflugene]{fluburdenbyflugene_nocoinfection}).
{\bf (C)}
The fraction of cells that express each of the four other genes among cells that express all RNP genes, as well as the fraction that express \emph{all} four of the other genes.
The fraction that express all four genes is well predicted by simply multiplying the frequencies of cells that express each gene individually, indicating that gene absence is approximately independent across these genes.
}

@@

@@

Cells that lack an RNP gene never derive more than a few percent of their mRNAs from virus, confirming the expected result that all four RNP genes are essential for high levels of viral transcription (\FIG{fluburdenbyflugene}A, \FIGSUPP[fluburdenbyflugene]{fluburdensamplebreakdown}, \FIGSUPP[fluburdenbyflugene]{fluburden_nocoinfection}).
However, we observe cells that lack each of the other non-RNP genes but still derive $\approx$40\% of their mRNAs from virus, suggesting that none of the other genes are important for high levels of viral transcription.
These results are statistically supported by \FIG{fluburdenbyflugene}B, which shows that absence of any RNP gene but \emph{not} any other viral gene is associated with reduced amounts of viral mRNA. 
However, gene absence clearly does not explain all of the variability in viral gene expression, since even cells expressing all viral genes exhibit a very wide distribution in the amount of viral mRNA that they express.
Specifically, at both 8 and 10 hours, the amount of viral mRNA in individual cells expressing all eight viral genes still ranges from $<$1\% to $>$50\% (\FIG{fluburdenbyflugene}A, \FIGSUPP[fluburdenbyflugene]{fluburdensamplebreakdown}, \FIGSUPP[fluburdenbyflugene]{fluburden_nocoinfection}).

>>>>

\begin{figure}[t!]
\centerline{\includegraphics[width=0.83\linewidth]{figures/p_flu_burden_flu_gene_merge.pdf}}
\caption{\label{fig:fluburdenbyflugene}
The absence of viral genes explains some of the variability in the amount of viral mRNA per cell.
{\bf (A)} 
The normalized expression of each viral gene as a function of the total fraction of mRNA in each infected cell derived from virus, taken over all time points.
Cells with high viral burden always express all RNP genes, but some cells with high viral burden lack each of the other genes.
Plots for individual samples are in \FIGSUPP[fluburdenbyflugene]{fluburdensamplebreakdown}, and a plot that excludes known coinfected cells is in \FIGSUPP[fluburdenbyflugene]{fluburden_nocoinfection}.
{\bf (B)}
Box and whisker plots showing the per-cell viral burden among cells with $>$0.5\% of their mRNA from virus, binned by whether or not the cells express each gene.
A Wilcoxon signed-rank test was used to test the null hypothesis that absence of each gene does not affect viral burden: **** = $P < 10^{-4}$, *** = $P < 10^{-3}$,  * = $P < 0.05$, ns = not significant.
The trends are similar if we look only at the 10-hour sample (\FIGSUPP[fluburdenbyflugene]{10hrfluburdenbyflugene}) or exclude known co-infected cells (\FIGSUPP[fluburdenbyflugene]{fluburdenbyflugene_nocoinfection}).
{\bf (C)}
The fraction of cells that express each of the four other genes among cells that express all RNP genes, as well as the fraction that express \emph{all} four of the other genes.
The fraction that express all four genes is well predicted by simply multiplying the frequencies of cells that express each gene individually, indicating that gene absence is approximately independent across these genes.
}

@@

\figdata{\label{figdata:simplemodel} Simulation with a simple model for the expected heterogeneity due to Poisson co-infection and presence / absence of the full RNP is in \texttt{simple\_Poisson\_model.html}.}

@@

Cells that lack an RNP gene never derive more than a few percent of their mRNAs from virus, confirming the expected result that all four RNP genes are essential for high levels of viral transcription (\FIG{fluburdenbyflugene}A, \FIGSUPP[fluburdenbyflugene]{fluburdensamplebreakdown}, \FIGSUPP[fluburdenbyflugene]{fluburden_nocoinfection}).
However, we observe cells that lack each of the other non-RNP genes but still derive $\approx$40\% of their mRNAs from virus, suggesting that none of the other genes are important for high levels of viral transcription.
These results are statistically supported by \FIG{fluburdenbyflugene}B, which shows that absence of any RNP gene but \emph{not} any other viral gene is associated with reduced amounts of viral mRNA. 
However, gene absence clearly does not explain all of the variability in viral gene expression, since even cells expressing all viral genes exhibit a very wide distribution in the amount of viral mRNA that they express.
Specifically, at both 8 and 10 hours, the amount of viral mRNA in individual cells expressing all eight viral genes still ranges from $<$1\% to $>$50\% (\FIG{fluburdenbyflugene}A, \FIGSUPP[fluburdenbyflugene]{fluburdensamplebreakdown}, \FIGSUPP[fluburdenbyflugene]{fluburden_nocoinfection}).  
Furthermore, the actual distribution of viral mRNA per infected cell (\FIG{viralbarcodes}E) is does not match the mostly bi-modal shape expected under a simple model where RNP gene absence and Poisson co-infection are the only factors (\FIG{fluburdenbyflugene}--source~data~\ref{figdata:simplemodel}), indicating that there are additional sources of variability beyond whether cells have full complement of RNP genes.
